The document title

### 2021-03-16 manuscript/manuscript.Rmd
Change markup characters

A convenience sample of `r max(sample_size$pGRF_N)` adults (`r unname(sample_size_sex["M"])` males; $`r sample_descriptives$age_mean`\pm`r sample_descriptives$age_sd`$ years; $`r sample_descriptives$height_mean`\pm`r sample_descriptives$height_sd`$ cm; $`r sample_descriptives$body_mass_mean`\pm`r sample_descriptives$body_mass_sd`$ kg; $`r sample_descriptives$BMI_mean`\pm`r sample_descriptives$BMI_sd`$ $\mathrm{kg\cdot m^{-2}}$, $\overline{X}\pm SD$) with no self-reported neurological or musculoskeletal limitations was recruited. Height and body mass were assessed following standard procedures with a stadiometer and a digital scale, respectively. All participants were informed about the experiments purpose and protocol before giving written informed consent. The study was approved by the local Ethics Committee (CES 192-14).

@@

Data collection took place at Porto Biomechanics Laboratory (LABIOMEP-UP) on an instrumented treadmill with built-in FP (AMTI Corporation, Watertown, USA). Before starting the data collection, participants were given time to warm up on the treadmill. The protocol consisted of walking and running in 13 incremental speeds, with 0º inclination, from 2 to 14 $\mathrm{km\cdot h^{-1}}$ with 1 $\mathrm{km\cdot h^{-1}}$ increment. Walking and running speeds ranged from 2 to 6 $\mathrm{km\cdot h^{-1}}$ and from 7 to 14 $\mathrm{km\cdot h^{-1}}$, respectively. Each walking speed lasted 1 minute and each running speed 30 seconds. To ensure single foot contact on each FP during the trial, feedback regarding location on the treadmill was provided by a researcher. Participants wore their own sports shoes throughout the trials.

@@

Activity monitors used were the GT9X Link model ($\mathrm{\pm16}g$ range; ActiGraph, Pensacola, USA) that incorporates primary and secondary triaxial accelerometers. As a manufacturer proprietary filter is applied on the primary, but not on the secondary accelerometer raw data, and therefore only data from the latter was used in this study, reducing filtering bias and enhancing the data processing replicability.

@@

Data from FP and accelerometers were collected at 1000 Hz and 100 Hz sampling frequency, respectively, and operated through the manufacturer supplied software (for the FP: Netforce, Version 3.5.1; AMTI Corporation, Watertown, USA; for the accelerometers: ActiLife version 6.13.3; Actigraph, Pensacola, USA). Raw GRF (expressed in N) and ACC (expressed in gravitational ACC units; 1$g$ = 9.807 $\mathrm{m\cdot s^{-2}}$) from the x, y and z vectors were exported to plain text files.

@@

Regression equations to predict pGFR and pLR in each vector (resultant and vertical) for each accelerometer placement (ankle, lower back and hip) were developed by linear mixed models. Body mass (kg) was tested as a predictor in all models. Additionally, pACC and pAR were tested as predictors in the pGRF and pLR models, respectively. All predictors were tested as fixed effects and were shown to be significant. Subject and speed were tested as random effects, and the inclusion of both has showed to improve the models. Final models were chosen according to -2 log-likelihood statistics [@Field2012]. Traditional coefficient of determination ($\mathrm{R^2}$) was represented by conditional $\mathrm{R^2}$, that estimates the variance explained by the whole model [@Nakagawa2013].

@@

To assess the models prediction accuracy, mean absolute error (MAE), mean absolute percent error (MAPE) and root mean square error (RMSE) were calculated. Also, methods agreement was evaluated with Bland-Altman plots by plotting the difference between the actual (FP measured) and predicted values against their mean. Bias was computed as the mean of these differences and the limits of agreement were obtained using \pm 1.96 standard deviation of the mean between actual and predicted values [@Bland1986]. In order to test whether there was a consistent over or underestimation of the acual values, one-sample \textit{t} tests were performed to assess if the difference between actual and predicted values are different than zero. Linear regressions were applied to identify if the magnitude of the mean between actual and predicted values influenced the magnitude of their difference. This was done to ascertain if the prediction accuracy was constant throughout the assayed magnitude range, which is known as proportional bias [@Giavarina2015]. These analyses were conducted separately for data from each accelerometer placement and outcome.

@@

Our recruited sample presented a broad range of body masses, from normal weight to severe obesity. It would be expected that participants with higher BMI values, due to low motor ability and physical capacity, are able to perform the trials only at the lowest speeds. This was precisely what happened and, as can be observed in Figure 1, from 2 to 5 $\mathrm{km\cdot h^{-1}}$ the BMI median was around 30 $\mathrm{kg\cdot m^{-2}}$, decreasing to around 27 $\mathrm{kg\cdot m^{-2}}$ at 6 $\mathrm{km\cdot h^{-1}}$, which was the fastest walking speed, with a further reduction to less than 25 $\mathrm{kg\cdot m^{-2}}$ at the running speeds. Furthermore, only one participant with a BMI higher than 30 $\mathrm{kg\cdot m^{-2}}$ was able to run. This implies in a sample size reduction from `r filter(sample_size, acc_placement == "hip" & vector == "resultant" & speed == "2")$pGRF_N` at the slowest speed to `r filter(sample_size, acc_placement == "hip" & vector == "resultant" & speed == "14")$pGRF_N` at the fastest.

@@

Regression equations were developed to predict the mechanical loading variables based on accelerometry data. In the pGRF prediction equations, body mass and pACC were significant predictors and included in the models. Table 1 shows the regression coefficients, $\mathrm{R^2}$ and accuracy indices from models derived from accelerometers placed at ankle, lower back and hip, and from the resultant vector and its vertical component. In all pGRF prediction models, $\mathrm{R^2}$ values were `r max(GRF_accuracy$R2_cond)`, which shows that the predictor variables are able to explain the variability in the outcome varibale to a great extent. Also, MAE ranged between `r myround(min(GRF_accuracy$MAE), 1)` N and `r myround(max(GRF_accuracy$MAE), 1)` N, and the MAPE from `r myround(min(GRF_accuracy$MAPE) * 100, 1)`\% to `r myround(max(GRF_accuracy$MAPE) * 100, 1)`\%, which is a smaller error than previously published equations [@Neugebauer2014; @Fortune2014], and similar to the error from equations previously published by our group, but which included walking only [@Veras2020]. RMSE were also computed and shown to range between `r myround(min(GRF_accuracy$RMSE), 1)` N and `r myround(max(GRF_accuracy$RMSE), 1)` N. Of all the pGRF predictions equations developed, the one based on hip-worn accelerometers and the resultant vector has shown the best performance, with an $\mathrm{R^2}$ of `r hip_GRF_accuracy$R2_cond`, and the lowest MAE (`r myround(hip_GRF_accuracy$MAE, 1)` N), MAPE (`r myround(hip_GRF_accuracy$MAPE * 100, 1)`\%) and RMSE (`r myround(hip_GRF_accuracy$RMSE, 1)` N).

@@

Moreover, pLR prediction equations were developed, and these included as predictors the pAR and again body mass. Regression coefficients, $\mathrm{R^2}$ and accuracy indices for these equations are also shown in Table 1. For these equations, the lowest $\mathrm{R^2}$ found was `r myround(min(LR_accuracy$R2_cond), 2)` and the highest was `r myround(max(LR_accuracy$R2_cond), 2)`, reflecting a moderate to high prediction ability. MAE, MAPE and RMSE values ranged from `r myround(min(LR_accuracy$MAE), 1)` N to `r myround(max(LR_accuracy$MAE), 1)` N, from `r myround(min(LR_accuracy$MAPE) * 100, 1)`\% to `r myround(max(LR_accuracy$MAPE) * 100, 1)`\%, and from `r myround(min(LR_accuracy$RMSE), 1)` N to `r myround(max(LR_accuracy$RMSE), 1)` N, respectively. The results found for the pLR prediction equations were all considerably poorer than those for the pGRF equations. The accuracy indices values were also somewhat worse than previously published pLR prediction equations [@Veras2020; @Fortune2014], although equations from both studies were only validate for walking. The pLR equation with the lower prediction error was the one for accelerometers worn at lower back and the resultant vector, with an $\mathrm{R^2}$ of `r back_LR_accuracy$R2_cond` and a MAE, MAPE and RMSE of `r myround(back_LR_accuracy$MAE, 1)` N, `r myround(back_LR_accuracy$MAPE * 100, 1)`\% and `r myround(back_LR_accuracy$RMSE, 1)` N, respectively.

@@

Afterwards, Bland-Altman plots were built to visualize the agreement between the actual and predicted pGRF and pLR, for both the resultant vector and its vertical component from all accelerometer placements. Bland-Altman plots for the pGRF and pLR resultant vector from hip-worn accelerometers can be seen in Figure 3, and for all accelerometer placements and both resultant and vertical vectors in the Supplemental Figures S3 and S4. Values for all pGRF and pLR prediction equations developed tended to assemble around zero and between the limits of agreement. For the pGRF equations, a tendency to higher dispersion can be observed in the higher magnitudes, which correspond to the running trials, and especially in values derived from ankle-worn accelerometers (Supplemental Figure S3, Panels A and D). Prediction equations for pLR showed a more accentuated dispersion, and again mainly in the higher magnitudes (Supplemental Figure S4). The pLR prediction model based on data from accelerometers placed at the ankle and the resultant vector displayed a bias different than zero ($p = `r myround(ba_bias_test_res_LR$ankle["p.value"], 2)`$). None of the other models presented a bias different than zero ($p > 0.05$), demonstrating that they do not consistently over or underestimate the pGRF and pLR values. All equations based on ankle-worn accelerometer data, for both outcomes (pGRF and pLR) and vectors (resultant and vertical) showed a significant proportional bias ($p < 0.05$). Despite the statistical significance, linear regressions presented a low $\mathrm{R^2}$ (highest $\mathrm{R^2} = `r myround(max(ba_regression_ankle_R2), 2)`$).

>>>>

A convenience sample of `r max(sample_size$pGRF_N)` adults (`r unname(sample_size_sex["M"])` males; `r sample_descriptives$age_mean`±`r sample_descriptives$age_sd` years; `r sample_descriptives$height_mean`±`r sample_descriptives$height_sd` cm; `r sample_descriptives$body_mass_mean`±`r sample_descriptives$body_mass_sd` kg; `r sample_descriptives$BMI_mean`±`r sample_descriptives$BMI_sd` kg·m^-2^, X̄±SD) with no self-reported neurological or musculoskeletal limitations was recruited. Height and body mass were assessed following standard procedures with a stadiometer and a digital scale, respectively. All participants were informed about the experiments purpose and protocol before giving written informed consent. The study was approved by the local Ethics Committee (CES 192-14).

@@

Data collection took place at Porto Biomechanics Laboratory (LABIOMEP-UP) on an instrumented treadmill with built-in FP (AMTI Corporation, Watertown, USA). Before starting the data collection, participants were given time to warm up on the treadmill. The protocol consisted of walking and running in 13 incremental speeds, with 0º inclination, from 2 to 14 km·h^-1^ with 1 km·h^-1^ increment. Walking and running speeds ranged from 2 to 6 km·h^-1^ and from 7 to 14 km·h^-1^, respectively. Each walking speed lasted 1 minute and each running speed 30 seconds. To ensure single foot contact on each FP during the trial, feedback regarding location on the treadmill was provided by a researcher. Participants wore their own sports shoes throughout the trials.

@@

Activity monitors used were the GT9X Link model (±16*g* range; ActiGraph, Pensacola, USA) that incorporates primary and secondary triaxial accelerometers. As a manufacturer proprietary filter is applied on the primary, but not on the secondary accelerometer raw data, and therefore only data from the latter was used in this study, reducing filtering bias and enhancing the data processing replicability.

@@

Data from FP and accelerometers were collected at 1000 Hz and 100 Hz sampling frequency, respectively, and operated through the manufacturer supplied software (for the FP: Netforce, Version 3.5.1; AMTI Corporation, Watertown, USA; for the accelerometers: ActiLife version 6.13.3; Actigraph, Pensacola, USA). Raw GRF (expressed in N) and ACC (expressed in gravitational ACC units; 1*g* = 9.807 m·s^-2^) from the x, y and z vectors were exported to plain text files.

@@

Regression equations to predict pGFR and pLR in each vector (resultant and vertical) for each accelerometer placement (ankle, lower back and hip) were developed by linear mixed models. Body mass (kg) was tested as a predictor in all models. Additionally, pACC and pAR were tested as predictors in the pGRF and pLR models, respectively. All predictors were tested as fixed effects and were shown to be significant. Subject and speed were tested as random effects, and the inclusion of both has showed to improve the models. Final models were chosen according to -2 log-likelihood statistics [@Field2012]. Traditional coefficient of determination (R^2^) was represented by conditional R^2^, that estimates the variance explained by the whole model [@Nakagawa2013].

@@

To assess the models prediction accuracy, mean absolute error (MAE), mean absolute percent error (MAPE) and root mean square error (RMSE) were calculated. Also, methods agreement was evaluated with Bland-Altman plots by plotting the difference between the actual (FP measured) and predicted values against their mean. Bias was computed as the mean of these differences and the limits of agreement were obtained using ±1.96 standard deviation of the mean between actual and predicted values [@Bland1986]. In order to test whether there was a consistent over or underestimation of the acual values, one-sample \textit{t} tests were performed to assess if the difference between actual and predicted values are different than zero. Linear regressions were applied to identify if the magnitude of the mean between actual and predicted values influenced the magnitude of their difference. This was done to ascertain if the prediction accuracy was constant throughout the assayed magnitude range, which is known as proportional bias [@Giavarina2015]. These analyses were conducted separately for data from each accelerometer placement and outcome.

@@

Our recruited sample presented a broad range of body masses, from normal weight to severe obesity. It would be expected that participants with higher BMI values, due to low motor ability and physical capacity, are able to perform the trials only at the lowest speeds. This was precisely what happened and, as can be observed in Figure 1, from 2 to 5 km·h^-1^ the BMI median was around 30 kg·m^-2^, decreasing to around 27 kg·m^-2^ at 6 km·h^-1^, which was the fastest walking speed, with a further reduction to less than 25 kg·m^-2^ at the running speeds. Furthermore, only one participant with a BMI higher than 30 kg·m^-2^ was able to run. This implies in a sample size reduction from `r filter(sample_size, acc_placement == "hip" & vector == "resultant" & speed == "2")$pGRF_N` at the slowest speed to `r filter(sample_size, acc_placement == "hip" & vector == "resultant" & speed == "14")$pGRF_N` at the fastest.

@@

Regression equations were developed to predict the mechanical loading variables based on accelerometry data. In the pGRF prediction equations, body mass and pACC were significant predictors and included in the models. Table 1 shows the regression coefficients, R^2^ and accuracy indices from models derived from accelerometers placed at ankle, lower back and hip, and from the resultant vector and its vertical component. In all pGRF prediction models, R^2^ values were `r max(GRF_accuracy$R2_cond)`, which shows that the predictor variables are able to explain the variability in the outcome varibale to a great extent. Also, MAE ranged between `r myround(min(GRF_accuracy$MAE), 1)` N and `r myround(max(GRF_accuracy$MAE), 1)` N, and the MAPE from `r myround(min(GRF_accuracy$MAPE) * 100, 1)`\% to `r myround(max(GRF_accuracy$MAPE) * 100, 1)`\%, which is a smaller error than previously published equations [@Neugebauer2014; @Fortune2014], and similar to the error from equations previously published by our group, but which included walking only [@Veras2020]. RMSE were also computed and shown to range between `r myround(min(GRF_accuracy$RMSE), 1)` N and `r myround(max(GRF_accuracy$RMSE), 1)` N. Of all the pGRF predictions equations developed, the one based on hip-worn accelerometers and the resultant vector has shown the best performance, with an R^2^ of `r hip_GRF_accuracy$R2_cond`, and the lowest MAE (`r myround(hip_GRF_accuracy$MAE, 1)` N), MAPE (`r myround(hip_GRF_accuracy$MAPE * 100, 1)`\%) and RMSE (`r myround(hip_GRF_accuracy$RMSE, 1)` N).

@@

Moreover, pLR prediction equations were developed, and these included as predictors the pAR and again body mass. Regression coefficients, R^2^ and accuracy indices for these equations are also shown in Table 1. For these equations, the lowest R^2^ found was `r myround(min(LR_accuracy$R2_cond), 2)` and the highest was `r myround(max(LR_accuracy$R2_cond), 2)`, reflecting a moderate to high prediction ability. MAE, MAPE and RMSE values ranged from `r myround(min(LR_accuracy$MAE), 1)` N to `r myround(max(LR_accuracy$MAE), 1)` N, from `r myround(min(LR_accuracy$MAPE) * 100, 1)`\% to `r myround(max(LR_accuracy$MAPE) * 100, 1)`\%, and from `r myround(min(LR_accuracy$RMSE), 1)` N to `r myround(max(LR_accuracy$RMSE), 1)` N, respectively. The results found for the pLR prediction equations were all considerably poorer than those for the pGRF equations. The accuracy indices values were also somewhat worse than previously published pLR prediction equations [@Veras2020; @Fortune2014], although equations from both studies were only validate for walking. The pLR equation with the lower prediction error was the one for accelerometers worn at lower back and the resultant vector, with an R^2^ of `r back_LR_accuracy$R2_cond` and a MAE, MAPE and RMSE of `r myround(back_LR_accuracy$MAE, 1)` N, `r myround(back_LR_accuracy$MAPE * 100, 1)`\% and `r myround(back_LR_accuracy$RMSE, 1)` N, respectively.

@@

Afterwards, Bland-Altman plots were built to visualize the agreement between the actual and predicted pGRF and pLR, for both the resultant vector and its vertical component from all accelerometer placements. Bland-Altman plots for the pGRF and pLR resultant vector from hip-worn accelerometers can be seen in Figure 3, and for all accelerometer placements and both resultant and vertical vectors in the Supplemental Figures S3 and S4. Values for all pGRF and pLR prediction equations developed tended to assemble around zero and between the limits of agreement. For the pGRF equations, a tendency to higher dispersion can be observed in the higher magnitudes, which correspond to the running trials, and especially in values derived from ankle-worn accelerometers (Supplemental Figure S3, Panels A and D). Prediction equations for pLR showed a more accentuated dispersion, and again mainly in the higher magnitudes (Supplemental Figure S4). The pLR prediction model based on data from accelerometers placed at the ankle and the resultant vector displayed a bias different than zero (*p* = `r myround(ba_bias_test_res_LR$ankle["p.value"], 2)`). None of the other models presented a bias different than zero (*p* > 0.05), demonstrating that they do not consistently over or underestimate the pGRF and pLR values. All equations based on ankle-worn accelerometer data, for both outcomes (pGRF and pLR) and vectors (resultant and vertical) showed a significant proportional bias (*p* < 0.05). Despite the statistical significance, linear regressions presented a low R^2^ (highest R^2^ = `r myround(max(ba_regression_ankle_R2), 2)`).
